LOINC2HPO_EPIC AND SCC

### 2022-03-08 src/main/r/MSDW_HPO_Phenotype_Analysis.Rmd
update

@@

```{r}
plot_phenotypes_of_interest_by_year <- function(terms_of_interest, df){
  hpo_terms_selected <- df %>% filter(termid %in% terms_of_interest) %>% select(termid, label) %>% distinct()

  p <- df %>% filter(termid %in% terms_of_interest) %>% 
    filter(lab_year <= 2019 & lab_year > 2000) %>%
    ggplot(aes(x = lab_year, y = n_patient, group = termid, color = termid)) + 
    geom_point() +
    geom_line() +
    scale_color_discrete(breaks = hpo_terms_selected$termid, labels = str_c(hpo_terms_selected$termid, hpo_terms_selected$label, sep = " "))

@@

hpo_level_1_terms <- phenotype_rank_all_by_year %>% filter(distancetoroot == 1) %>% select(termid, label) %>% distinct()
plot_phenotypes_of_interest_by_year(terms_of_interest = hpo_level_1_terms$termid, df=phenotype_rank_all_by_year)
```

@@

  if (force_rerun | !file.exists(df_path)){
    df <- dbGetQuery(dsca, "
               with testable_phenotypes_direct as (
              	select lab.medical_record_number , lab.context_name , lab.lab_date , lab.lab_time , lab.loinc , testable.termid , testable.is_tested_directly 
              	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july lab
              	join hai_az_prod.loinc_testable_phenotypes testable on lab.loinc = testable.loincid 
              	--where testable.is_tested_directly 
              ),
              testable_phenotype_direct_freq as (
              	select termid, count(*) as n_record, count(distinct medical_record_number) as n_patient
              	from testable_phenotypes_direct
              	group by termid 
              )
              select testable.*, hpo.label, hpo.distancetoroot 
              from testable_phenotype_direct_freq testable
              join hai_az_prod.hpo hpo on testable.termid = hpo.termid
              order by n_record desc;
              ")

@@

Plot the anual change of tested phenotypes.

@@

```{r}
hpo_level_1_terms <- tested_phenotype_rank_all_by_year %>% filter(distancetoroot == 1) %>% select(termid, label) %>% distinct()
plot_phenotypes_of_interest_by_year(terms_of_interest = hpo_level_1_terms$termid, df=tested_phenotype_rank_all_by_year)
```

@@

```{r}
children_of_interest <- c('HP:0410367', 'HP:0031047', 'HP:0003237', 'HP:0003212', 'HP:0003261', 'HP:0410369', 'HP:0410371', 'HP:0410246', 'HP:0003496')
plot_phenotypes_of_interest_by_year(terms_of_interest = children_of_interest, df=tested_phenotype_rank_all_by_year)
```

>>>>

### Collect patient basic demographics
```{r}
get_patient_race_sex_distribution <- function(force_rerun=FALSE, cache_dir=CACHE_DIR){
  df_path = glue('{cache_dir}/patient_count_by_race_sex.csv')

  if (force_rerun | !file.exists(df_path)){
    df <- dbGetQuery(dsca, "
            select race, gender, count(*) as n
            from hai_az_prod.person_msdw_2020july pmj 
            group by race, gender 
            order by race, gender;
            ")

  return (df)

}

patient_race_sex_distribution <- get_patient_race_sex_distribution()

# recode all unknown gender as one category
patient_race_sex_distribution <- patient_race_sex_distribution %>% 
  mutate(gender = recode(gender, 'MSDW_NOT APPL' = 'Unknown')) %>%
  mutate(gender = recode(gender, 'MSDW_OTHER' = 'Unknown')) %>%
  mutate(gender = recode(gender, 'MSDW_UNKNOWN' = 'Unknown')) %>%
  mutate(gender = recode(gender, 'NOT AVAILABLE'= 'Unknown')) %>%
  mutate(gender = recode(gender, .missing = 'Unknown'))

# recode "African American" as "Black or African American"
# TODO: update the race map
patient_race_sex_distribution <- patient_race_sex_distribution %>%
  mutate(race = recode(race, 'African American' = 'Black or African American'))
```

```{r}
countByGender <- patient_race_sex_distribution %>% 
  group_by(gender) %>% 
  summarise(n = sum(n))

plot_by_sex <- ggplot(countByGender) + 
  geom_bar(aes(x = 1, y = n / sum(n), fill = gender), stat = "identity") + 
  scale_fill_manual(name = "Sex", values = c(Male = "royalblue", Female = "lightcoral", Indeterminant = 'Cyan', Unknown = 'Orange')) + 
  xlab("") + ylab("patient count") +
  coord_polar(theta = "y") + 
  theme_bw() + 
  theme(panel.grid = element_blank(), legend.position = c(0.25, 0.7), 
        axis.text = element_blank(), axis.ticks = element_blank(), 
        axis.title = element_text(size = 8), legend.text = element_text(size = 8), 
        legend.title = element_text(size = 10), 
        legend.background = element_rect(fill=alpha('white', 0.1)))

ggsave(filename = glue('{CACHE_DIR}/plot_by_sex.png'), plot = plot_by_sex, width = 2.5, height = 2.5)
plot_by_sex

```

```{r}
countByRace_Gender <- patient_race_sex_distribution

countByRace_Gender$race <- if_else(countByRace_Gender$race == 'Unknown', 'Unreported', countByRace_Gender$race)

plot_by_race  <- ggplot(countByRace_Gender %>% 
         filter(gender %in% c('Male', 'Female')) %>% 
         mutate(race = as.factor(race), n = if_else(gender == "Male", n, -n))) + 
  geom_bar(aes(x = reorder(race, -n, FUN=sum), y = n, fill = gender, group=gender), stat="identity") + 
##  scale_y_continuous(breaks = seq(-40,40, 10), labels = abs(seq(-40,40, 10))) +
  scale_fill_manual(name = "Sex", values = c(Male = "royalblue", Female = "lightcoral")) + 
  scale_y_continuous(labels = abs) +
  xlab("") + ylab("patient count") +
  coord_flip() + 
  theme_bw() + 
  theme(panel.grid = element_blank(), legend.position = "right", 
        axis.title = element_text(size = 8), axis.text = element_text(size = 8), 
        legend.text = element_text(size = 8), legend.title = element_text(size = 10), 
        legend.background = element_rect(fill=alpha('white', 0.1)))

ggsave(filename = glue("{CACHE_DIR}/plotByRace.png"), plot = plot_by_race, width = 5, height = 2)
plot_by_race
```

Patient age analysis: find the birth years for all the patients and find their current age
```{r}
get_patient_birth_year_distribution <- function(force_rerun=FALSE, cache_dir=CACHE_DIR){
  df_path = glue('{cache_dir}/patient_count_by_birth_year_sex.csv')

  if (force_rerun | !file.exists(df_path)){
    df <- dbGetQuery(dsca, "
            with person_birth_year as (
            	select medical_record_number , date_of_birth , gender , date_part('Year', date_of_birth) as birth_year 
            	from hai_az_prod.person_msdw_2020july pmj
            )
            select birth_year, gender, count(*) as n 
            from person_birth_year 
            group by birth_year , gender 
            order by birth_year, gender;
            ")

  return (df)

}

patient_birth_year_sex_distribution <- get_patient_birth_year_distribution()

# lump together sex that can be treated as Unknown
patient_birth_year_sex_distribution <- patient_birth_year_sex_distribution %>% 
  mutate(gender = recode(gender, 'MSDW_NOT APPL' = 'Unknown')) %>%
  mutate(gender = recode(gender, 'MSDW_OTHER' = 'Unknown')) %>%
  mutate(gender = recode(gender, 'MSDW_UNKNOWN' = 'Unknown')) %>%
  mutate(gender = recode(gender, 'NOT AVAILABLE'= 'Unknown')) %>%
  mutate(gender = recode(gender, .missing = 'Unknown')) %>%
  # sample_n(50) %>%
  # ignore ~ 1 million patients whoes birth_year is unknown
  # filter(!is.na(birth_year)) %>%
  # 2 patients have birth_year after 2020, which is clearly error as our database cut off is 2020
  filter(birth_year <= 2020) %>%
  mutate(birth_year_bucket = cut(birth_year, breaks = c(1800, seq(1931, 2021, by = 10)), dig.lab = 4)) %>%
  group_by(birth_year_bucket, gender) %>% 
  summarise(n = sum(n)) %>% 
  arrange(birth_year_bucket)

plot_by_birth_year_sex <- patient_birth_year_sex_distribution %>% ungroup() %>% 
  filter(gender %in% c('Male', 'Female')) %>% 
  mutate(n = if_else(gender == 'Female', -n, n)) %>%
  mutate(birth_year_bucket = factor(birth_year_bucket, levels = rev(levels(birth_year_bucket)))) %>% 
  # change bucket name (1800,1931] to '<= 1931'
  # change bucket name (2011, 2021] to '(2011, 2020]' -- our database cutoff is 2020
  mutate(birth_year_bucket = recode(birth_year_bucket, '(1800,1931]' = '<=1931')) %>% 
  mutate(birth_year_bucket = recode(birth_year_bucket, '(2011,2021]' = '(2011,2020]')) %>%{
    ggplot(.) +
  geom_bar(aes(x = birth_year_bucket, y = n, fill = gender), stat = 'identity') + 
  scale_fill_manual(name = "Sex", values = c(Male = "royalblue", Female = "lightcoral")) + 
  scale_y_continuous(labels = abs) +
  xlab('') + ylab('patient count') +
  coord_flip() + 
      theme_bw() + 
  theme(panel.grid = element_blank(), legend.position = "right", 
        axis.title = element_text(size = 8), axis.text = element_text(size = 8), 
        legend.text = element_text(size = 8), legend.title = element_text(size = 10), 
        legend.background = element_rect(fill=alpha('white', 0.1)))
  }

ggsave(filename = glue("{CACHE_DIR}/plotByBirthyearSex.png"), plot = plot_by_birth_year_sex, width = 4, height = 3)
plot_by_birth_year_sex

```

### Count patient distribution by the number of HPO terms per patient 
```{r}

```

### Collect patient counts for those with lab data at each year

```{r}
get_tested_patient_count_by_year <- function(force_rerun=FALSE, cache_dir=CACHE_DIR){
  df_path = glue('{cache_dir}/tested_patient_count_by_year.csv')

  if (force_rerun | !file.exists(df_path)){
    df <- dbGetQuery(dsca, "
            with scc_lab as (
            	select medical_record_number , calendar_date as lab_date, date_part('Year', lab_date) as lab_year
            	from hai_az_prod.lab_scc_2020july lsj
            ), 
            epic_lab as (
            	select mrn as medical_record_number, order_date as lab_date, date_part('Year', lab_date) as lab_year
            	from hai_az_test.lab_epic_2020july lej
            ),
            scc_and_epic as (
            	select * 
            	from scc_lab 
            	union all
            	select * 
            	from epic_lab
            )
            select lab_year, count(distinct medical_record_number) as n_patient
            from scc_and_epic 
            group by lab_year
            order by lab_year;
            ")

tested_patient_count_by_year <- get_tested_patient_count_by_year()
```

@@

```{r}
plot_phenotypes_of_interest_by_year <- function(terms_of_interest, df, normalize=TRUE){
  hpo_terms_selected <- df %>% filter(termid %in% terms_of_interest) %>% select(termid, label) %>% distinct()

  y_str = ifelse(normalize, 'n_patient/n_tested', 'n_patient')
  y_label = ifelse(normalize, 'patient fraction among tested', 'patient count')

  p <- df %>% filter(termid %in% terms_of_interest) %>% 
    filter(lab_year <= 2019 & lab_year > 2000) %>%
    ggplot(aes(x = lab_year, y = eval(parse(text=y_str)), group = termid, color = termid)) + 
    geom_point() +
    geom_line() +
    scale_color_discrete(breaks = hpo_terms_selected$termid, labels = str_c(hpo_terms_selected$termid, hpo_terms_selected$label, sep = " ")) +
    ylab(y_label)

@@

hpo_level_1_terms <- phenotype_rank_all_by_year %>% filter(distancetoroot == 1) %>% select(termid, label) %>% distinct()
plot_phenotypes_of_interest_by_year(terms_of_interest = hpo_level_1_terms$termid, df=phenotype_rank_all_by_year, normalize = FALSE)
```

```{r}
hpo_level_1_terms <- phenotype_rank_all_by_year %>% filter(distancetoroot == 1) %>% select(termid, label) %>% distinct()
plot_phenotypes_of_interest_by_year(terms_of_interest = hpo_level_1_terms$termid, df= phenotype_rank_all_by_year %>% left_join(tested_patient_count_by_year %>% rename(n_tested = n_patient), by = 'lab_year'), normalize = TRUE)
```

@@

  if (force_rerun | !file.exists(df_path)){
    df <- dbGetQuery(dsca, "
               with testable_phenotypes_all as (
              	select lab.medical_record_number , lab.context_name , lab.lab_date , lab.lab_time , lab.loinc , testable.termid , testable.is_tested_directly 
              	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july lab
              	join hai_az_prod.loinc_testable_phenotypes testable on lab.loinc = testable.loincid 
              	--where testable.is_tested_directly 
              ),
              testable_phenotype_all_freq as (
              	select termid, count(*) as n_record, count(distinct medical_record_number) as n_patient
              	from testable_phenotypes_all
              	group by termid 
              )
              select testable.*, hpo.label, hpo.distancetoroot 
              from testable_phenotype_all_freq testable
              join hai_az_prod.hpo hpo on testable.termid = hpo.termid
              order by n_record desc;
              ")

@@

Plot the annual change of tested phenotypes.

@@

```{r}
hpo_level_1_terms <- tested_phenotype_rank_all_by_year %>% filter(distancetoroot == 1) %>% select(termid, label) %>% distinct()
plot_phenotypes_of_interest_by_year(terms_of_interest = hpo_level_1_terms$termid, df=tested_phenotype_rank_all_by_year, normalize = FALSE)
```
```{r}
plot_phenotypes_of_interest_by_year(terms_of_interest = hpo_level_1_terms$termid, df=tested_phenotype_rank_all_by_year %>% left_join(tested_patient_count_by_year %>% rename(n_tested = n_patient), by = 'lab_year'), normalize = TRUE)
```

@@

Find what local lab codes are mapped to HP:0010702 Increased circulating antibody level and find which are tested more often after 2017.
```{r}
get_tested_phenotype_spike_by_loinc_year <- function(force_rerun=FALSE, cache_dir=CACHE_DIR){
  df_path = glue('{cache_dir}/tested_phenotype_spike_by_loinc_year.csv')

  if (force_rerun | !file.exists(df_path)){
    df <-  dbGetQuery(dsca, "
              with lab_spike as (
            	select context_name , lab.loinc , ltp.termid , date_part('Year', lab_date) as lab_year, count(*) as n_record, count(distinct medical_record_number) as n_patient
            	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july lab 
            	join hai_az_prod.loinc_testable_phenotypes ltp on lab.loinc = ltp.loincid 
            	where ltp.termid = 'HP:0010702'
            	group by context_name , lab.loinc , ltp.termid, date_part('Year', lab_date)
            )
            select l.*
            from lab_spike l
            order by context_name , loinc , lab_year;
              ")

get_tested_phenotype_spike_by_localcode_year <- function(force_rerun=FALSE, cache_dir=CACHE_DIR){
  df_path = glue('{cache_dir}/tested_phenotype_spike_by_localcode_year.csv')

  if (force_rerun | !file.exists(df_path)){
    df <-  dbGetQuery(dsca, "
              with lab_spike as (
            	select context_name , local_test_code , ltp.termid, date_part('Year', lab_date) as lab_year, count(*) as n_record, count(distinct medical_record_number) as n_patient
            	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july lab 
            	join hai_az_prod.loinc_testable_phenotypes ltp on lab.loinc = ltp.loincid 
            	where ltp.termid = 'HP:0010702'
            	group by context_name , local_test_code ,ltp.termid,  date_part('Year', lab_date)
            ),
            local_lab_names as (
            	select distinct source, code, test_name
            	from hai_az_prod.loinc_mapping lm 
            )
            select l.*, r.test_name
            from lab_spike l
            join local_lab_names r on l.local_test_code = r.code
            order by context_name , local_test_code , lab_year;
              ")

tested_phenotype_spike_by_loinc_year <- get_tested_phenotype_spike_by_loinc_year()
tested_phenotype_spike_by_localcode_year <- get_tested_phenotype_spike_by_localcode_year()
```

We can see the spike is caused by LOINC:10834-0 Globulin [Mass/volume] in Serum by calculation.

The SCC spike is for LOINC:2465-3 IgG [Mass/volume] in Serum or Plasma. The increase is significant, but the raw count is still negligible. 
```{r}
tested_phenotype_spike_by_loinc_year %>% 
  left_join(tested_patient_count_by_year %>% rename(n_tested = n_patient), by = 'lab_year') %>%
    filter(lab_year <= 2019 & lab_year > 2000) %>%
  #filter(loinc == '10834-0') %>%
    ggplot(aes(x = lab_year, y = n_patient/n_tested, group = loinc, color = loinc)) + 
    geom_point() +
    geom_line() +
    #scale_color_discrete(breaks = loinc, labels = str_c('LOINC', loinc, sep = ":")) +
    ylab('patient fraction among tested') +
  facet_wrap(~context_name, nrow = 2, scales = 'free_y')
```

Similarly, if we use local codes for the plot, we can also see that EPIC lab test 3277 caused the spike, which has a name of "GLOBULIN". 
```{r}
tested_phenotype_spike_by_localcode_year %>% 
  left_join(tested_patient_count_by_year %>% rename(n_tested = n_patient), by = 'lab_year') %>%
    filter(lab_year <= 2019 & lab_year > 2000) %>%
  group_by(context_name, local_test_code) %>%
  mutate(min_year = min(lab_year), max_year = max(lab_year), max_min_year = max_year - min_year) %>%
  filter(max_min_year > 2) %>%
    ggplot(aes(x = lab_year, y = n_patient/n_tested, group = local_test_code, color = local_test_code)) + 
    geom_point() +
    geom_line() +
    #scale_color_discrete(breaks = loinc, labels = str_c('LOINC', loinc, sep = ":")) +
    ylab('patient fraction among tested') +
  theme(legend.position = 'bottom') +
  facet_wrap(~context_name, nrow = 2, scales = 'free_y')
```
